# Interactivity <i class="fas fa-mouse-pointer"></i>

Read: *IDVW2*, Chapter 10 Interactivity

<script src="https://d3js.org/d3.v7.js"></script>

## Lecture slides <i class="fa fa-television"></i>

[interactivity.pdf](pdfs/interactivity.pdf){target="_blank"}

## Binding event listeners

```{r, echo=FALSE, eval=!HTML, out.width='33%'}
webshot::webshot("https://jtr13.github.io/d3book/interactivity.html", selector = "svg")
```

```{asis, echo=HTML}
<svg width="300" height="200">
  <rect width="300" height="200" fill="lightblue"></rect>
  <circle cx="50" cy="75" r="20" fill="blue"></circle>
  <ellipse cx="175" cy="100" rx="45" ry="30" fill="green"></ellipse>
  <text x="100" y="150">(100, 150)</text>
  <line x1="250" y1="150" x2="275" y2="175" stroke="red" stroke-width="5"></line>
</svg>
```  

It's helpful to think carefully about what you want to happen when an event listener is triggered and what information you need. Open Developer Tools and try these in the Console. Note that event management [changed in v6](https://observablehq.com/@d3/d3v6-migration-guide#events){target="_blank"} so code written for earlier versions of D3 will not work. 


### Do something unrelated to the element that received the event

``` js
d3.select("svg")
  .on("click", function () {
    d3.select("svg")
      .append("text")
        .attr("x", "100")
        .attr("y", "40")
        .text("Hello World");
        });
```

### Change an attribute of the element that received the event

``` js
d3.select("line")
  .on("click", function() {
    d3.select(this)
      .attr("stroke-width", "10");
      });
```

In the context of event handlers, "this" is the element that received the event, a.k.a. what you clicked on if it's a click event. An alternative ($\geq$ v6, see link above) is to pass the event and access the element with `event.currentTarget`:


or

``` js
d3.select("line")
  .on("click", function(event) {
    d3.select(event.currentTarget)
      .attr("stroke", "yellow");
      });
```

### Get the value of an attribute of the element that received the event

``` js
d3.select("circle")
  .on("click", function(event) {
  var rad = d3.select(event.currentTarget).attr("r");
  d3.select("text")
    .text(`The radius is ${rad} pixels.`);
    });
```  

### Do something with the data bound to the element that received the event

``` js
d3.select("circle")
  .datum("4")
  .on("click", function(event, d) {
    d3.select(event.currentTarget) 
      .attr("stroke", "red")
      .attr("stroke-width", d => d);
      });
```

As in the previous example, `d3.select(this)` can be used instead of `d3.select(event.currentTarget)`.

Try changing the data value bound to the circle with `d3.select("circle").datum("10")` and clicking again.


### Get the svg location of the event

``` js
d3.select("svg")
  .on("click", function(event) {
    d3.select("text")
      .text(`(${d3.pointer(event).map(Math.round)})`)
      });
```

(Up to v5, `d3.mouse(this)` was used instead of `d3.pointer(event)`.)

### Do something with the value of a radio button

<p id="color" style="background-color: silver; color: white;">Please select your favorite primary color:</p>
<input type="radio" id="html" name="fav_color" value="red"> red
<input type="radio" id="css" name="fav_color" value="blue"> blue
<input type="radio" id="javascript" name="fav_color" value="yellow"> yellow

``` js
d3.selectAll("input")
  .on("click", function(event) {
    var favcolor = event.currentTarget.value;
    d3.select("p#color").style("color", favcolor);
    });
```    
<script>
d3.selectAll("input")
  .on("click", function(event) {
    var favcolor = event.currentTarget.value;
    d3.select("p#color").style("color", favcolor);
    });
</script>	

## Separating the function and event listener

Examples

``` js
function goyellow() {
  d3.select(this)
    .attr("fill", "yellow")
    };
```

``` js
d3.select("circle")
  .on("mouseover", goyellow);
```

Try this in the Console:  

``` js
d3.select("svg")
  .on("click", function (event) {
    console.log(d3.pointer(event).map(Math.round));
    });
``` 

### Attribute of an element

`d3.select(this).attr("id");`

### Value of radio button

`d3.select(this).node().value;`  (string)

`+d3.select(this).node().value;` (number)

## Add / remove "buttons"

(HTML paragraphs are used as buttons in this example.)

HTML:

``` html
<p id="add">Add an element</p>
<p id="remove_left">Remove bar (left)</p>
<p id="remove_right">Remove bar (right)</p>
```

JavaScript:

``` js
d3.selectAll("p")
    .on("click", function () {
      var paraID = d3.select(this).attr("id");
      if (paraID == "add") {
          var newvalue = Math.floor(Math.random()*400);
          bardata.push(newvalue);
          } else if (paraID == "remove_left") {
          bardata.shift();
          } else {
          bardata.pop();
          };
      update(bardata);
      });
```

## Putting it all together

Vertical bar chart with add / remove buttons and general update pattern

[vertical_bar.html](code/vertical_bar.html){target="_blank"}

## Dependent event listeners

In these examples, the behavior or existence of one event listener depends on another.

### Global variable example

Here the circle click behavior depends on the value of the radio button: if the "Move left" radio button is checked, the circle will move left *when clicked*. If the "Move right" radio button is checked, the circle will move right *when clicked*.

A global variable is used to keep track of the radio button value. The event listener on the circle conditions the behavior on the value of this global variable.

```{r, echo=FALSE, eval=!HTML, out.width='33%'}
webshot::webshot("https://jtr13.github.io/d3book/interactivity.html#global-variable-example", selector = "#rad")
```

```{asis, echo=HTML}
<div id="rad" style="margin-left: 30px">
<h4>Click the circle.</h4>
<input type="radio" name="direction" value="left" checked="true">&nbsp;Move left
<input type="radio" name="direction" value="right">&nbsp;Move right<br>
<svg id='radio' width='300' height='200'>
  <rect x='0' y='0' width='300' height ='200' fill = 'lightblue'></rect>
  <circle cx='150' cy='100' r='20' fill='red'></circle>
  <text x='10' y='190' style = 'font-size: 80%;'>svg#radio</text>
</svg>
</div>  
```
  

<div style="margin-left: 30px">
```{js}
// global variable keeps track of which radio button is clicked
var action = "left";
d3.select("div#rad")
  .selectAll("input")
  .on("click", function() { action = d3.select(this).node().value; });
	  
// circle click behavior depends on value of "action"
d3.select("svg#radio").select("circle")
  .on("click", function () {
    if (action == "left") {
      var cx_new = +d3.select(this).attr("cx") - 50;
      if (cx_new < 20) cx_new = 20;
      } else {
      var cx_new = +d3.select(this).attr("cx") + 50;
      if (cx_new > 280) cx_new = 280;
      }
    d3.select(this)
      .transition()
      .duration(500)
      .attr("cx", cx_new);
      });

```
</div>


### Turn off event listener

In this example, the event listeners on the squares are turned on or off depending on the value of the radio button. Event listeners can be removed by setting the behavior to `null`.

```{r, echo=FALSE, eval=!HTML, out.width='33%'}
webshot::webshot("https://jtr13.github.io/d3book/interactivity.html#turn-off-event-listener", selector = "#rad2")
```

```{asis, echo=knitr::is_html_output()}
<div id="rad2" style="margin-left: 30px">
<h4>Click a square.</h4>
<input type="radio" name="square" value="red" checked="true">&nbsp;Red active
<input type="radio" name="square" value="blue">&nbsp;Blue active<br>

<svg id='radio2' width='300' height='200'>
  <rect x='0' y='0' width='300' height ='200' fill = 'lightblue'></rect>
  <rect id='red' x='75' y='75' width='50' height='50' fill='red'></rect>
  <rect id='blue' x='175' y='75' width='50' height='50' fill='blue'></rect>
  <text x='10' y='190' style = 'font-size: 80%;'>svg#radio2</text>
</svg>
</div>
```

<div style="margin-left: 30px">

```{js}
// movement function
var jump = function () {
      d3.select(this).transition().duration(500)
      .attr('y', '0')
      .transition().duration(500).ease(d3.easeBounce)
      .attr('y', '75');
};

// initial setup: add event listener to red square
d3.select("svg#radio2")
  .select("rect#red")
  .on("click", jump);
    
// switch event listeners if radio button is clicked
d3.select("div#rad2").selectAll("input")
  .on("click", function () {
  if (d3.select(this).node().value == "blue") {
    d3.select("svg#radio2").select("rect#blue").on("click", jump);
    d3.select("svg#radio2").select("rect#red").on("click", null);
    } else {
    d3.select("svg#radio2").select("rect#red").on("click", jump);
    d3.select("svg#radio2").select("rect#blue").on("click", null);
    }
});

```

</div>
