# Interactivity <i class="fas fa-mouse-pointer"></i>

Read: *IDVW2*, Chapter 10 Interactivity

<script src="https://d3js.org/d3.v7.js"></script>

## Lecture slides <i class="fa fa-television"></i>

[interactivity.pdf](pdfs/interactivity.pdf){target="_blank"}

## Binding event listeners

```{r, echo=FALSE, eval=!HTML, out.width='33%'}
webshot::webshot("https://jtr13.github.io/d3book/interactivity.html", selector = "svg")
```

```{asis, echo=HTML}
<svg width="400" height="300">
  <rect width="300" height="200" fill="lightblue"></rect>
  <circle cx="50" cy="75" r="20" fill="blue"></circle>
  <ellipse cx="175" cy="100" rx="45" ry="30" fill="green"></ellipse>
  <text x="100" y="150">(100, 150)</text>
  <line x1="250" y1="150" x2="275" y2="175" stroke="red" stroke-width="5"></line>
</svg>
```  

In this chapter, we will explore some basic but interesting interactivities. When going through the section, read the script first before executing it in the console. For the following scripts open Developer Tools and try these in the Console. Think beforehead what will happen when an event listener is triggered.

Note that event management [changed in v6](https://observablehq.com/@d3/d3v6-migration-guide#events){target="_blank"} so code written for earlier versions of D3 will not work. 

### Add an element


``` js
d3.select("svg")
  .on("click", function () {
    d3.select("svg")
      .append("text")
        .attr("x", "100")
        .attr("y", "40")
        .text("Hello World");
        });
```

In this case, the insertion is irrelevant to the element that recevied the event.

### Change attribute of an element

``` js
d3.select("line")
  .on("click", function() {
    d3.select(this)
      .attr("stroke-width", "10");
      });
```

In the context of event handlers, "this" is the element that received the event, a.k.a. what you clicked on if it's a click event (In this case the red line). An alternative ($\geq$ v6, see link above) is to pass the event and access the element with `event.currentTarget`:

``` js
d3.select("line")
  .on("click", function(event) {
    d3.select(event.currentTarget)
      .attr("stroke", "yellow");
      });
```

### Get attribute value of an element

``` js
d3.select("circle")
  .on("click", function(event) {
  var rad = d3.select(event.currentTarget).attr("r");
  d3.select("text")
    .text(`The radius is ${rad} pixels.`);
    });
```  

In this example, we retrieve the radius of the circle after clicking on it.

### Change attribute using bounded data

``` js
d3.select("circle")
  .data([{s: "red", sw: "15"}])
  .on("click", function(event, d) {
    d3.select(event.currentTarget) 
      .attr("stroke", d.s)
      .attr("stroke-width", d.sw);
      });
```

**Note that starting with v6, the data is the 2nd parameter to be passed: `function(event, d)`. In addition, note that you do not need to pass `d` again when accessing the data: for example we use `d.s` not `d => d.s`**.

As in the previous example, `d3.select(this)` can be used instead of `d3.select(event.currentTarget)`.

Try changing the data value bound to the circle with `d3.select("circle").datum("10")` and clicking again.


### Get the svg location of the event

Run the following two scripts in the console.

``` js
d3.select("svg")
  .on("click", function(event) {
    d3.select("text")
      .text(`(${d3.pointer(event).map(Math.round)})`)
      });
```

``` js
d3.select("svg")
  .on("click", function (event) {
    console.log(d3.pointer(event).map(Math.round));
    });
``` 

In the first script, we get the location of a click and show it through the text field. In the second script, the location of a click is shown in the console.

### Change the value of a radio button

<p id="color" style="background-color: silver; color: white;">Please select your favorite primary color:</p>
<input type="radio" id="html" name="fav_color" value="red"> red
<input type="radio" id="css" name="fav_color" value="blue"> blue
<input type="radio" id="javascript" name="fav_color" value="yellow"> yellow

``` js
d3.selectAll("input")
  .on("click", function(event) {
    var favcolor = event.currentTarget.value;
    d3.select("p#color").style("color", favcolor);
    });
```  

<script>
d3.selectAll("input")
  .on("click", function(event) {
    var favcolor = event.currentTarget.value;
    d3.select("p#color").style("color", favcolor);
    });
</script>	

### Separating the function and event listener

We can also write a function and event listener separately to improve readability. For example, try the following in the console:

``` js
function goyellow() {
  d3.select(this)
    .attr("fill", "yellow")
    };
```

``` js
d3.select("circle")
  .on("mouseover", goyellow);
```

## Exercise: <i class="fas fa-dumbbell"></i>: functions

In this exercise, we will create a vertical bar chart with "buttions" to add or remove bars.

You can start with a vertical bar plot with general update patterns.

Hint: use HTML paragraphs as buttons

``` html
<p id="add">Add an element</p>
<p id="remove_left">Remove bar (left)</p>
<p id="remove_right">Remove bar (right)</p>
```

[solution](code/vertical_bar.html){target="_blank"}


## Dependent event listeners

In the following examples, the behavior or existence of one event listener depends on another.

### Global variable example

Here the circle click behavior depends on the value of the radio button: if the "Move left" radio button is checked, the circle will move left *when clicked*. If the "Move right" radio button is checked, the circle will move right *when clicked*.

A global variable is used to keep track of the radio button value. The event listener on the circle conditions the behavior on the value of this global variable.

```{r, echo=FALSE, eval=!HTML, out.width='33%'}
webshot::webshot("https://jtr13.github.io/d3book/interactivity.html#global-variable-example", selector = "#rad")
```

```{asis, echo=HTML}
<div id="rad" style="margin-left: 30px">
<h4>Click the circle.</h4>
<input type="radio" name="direction" value="left" checked="true">&nbsp;Move left
<input type="radio" name="direction" value="right">&nbsp;Move right<br>
<svg id='radio' width='300' height='200'>
  <rect x='0' y='0' width='300' height ='200' fill = 'lightblue'></rect>
  <circle cx='150' cy='100' r='20' fill='red'></circle>
  <text x='10' y='190' style = 'font-size: 80%;'>svg#radio</text>
</svg>
</div>  
```
  

<div style="margin-left: 30px">
```{js}
// global variable keeps track of which radio button is clicked
var action = "left";
d3.select("div#rad")
  .selectAll("input")
  .on("click", function() { action = d3.select(this).node().value; });
	  
// circle click behavior depends on value of "action"
d3.select("svg#radio").select("circle")
  .on("click", function () {
    if (action == "left") {
      var cx_new = +d3.select(this).attr("cx") - 50;
      if (cx_new < 20) cx_new = 20;
      } else {
      var cx_new = +d3.select(this).attr("cx") + 50;
      if (cx_new > 280) cx_new = 280;
      }
    d3.select(this)
      .transition()
      .duration(500)
      .attr("cx", cx_new);
      });

```
</div>


### Turn off event listener

In this example, the event listeners on the squares are turned on or off depending on the value of the radio button. Event listeners can be removed by setting the behavior to `null`.

```{r, echo=FALSE, eval=!HTML, out.width='33%'}
webshot::webshot("https://jtr13.github.io/d3book/interactivity.html#turn-off-event-listener", selector = "#rad2")
```

```{asis, echo=knitr::is_html_output()}
<div id="rad2" style="margin-left: 30px">
<h4>Click a square.</h4>
<input type="radio" name="square" value="red" checked="true">&nbsp;Red active
<input type="radio" name="square" value="blue">&nbsp;Blue active<br>

<svg id='radio2' width='300' height='200'>
  <rect x='0' y='0' width='300' height ='200' fill = 'lightblue'></rect>
  <rect id='red' x='75' y='75' width='50' height='50' fill='red'></rect>
  <rect id='blue' x='175' y='75' width='50' height='50' fill='blue'></rect>
  <text x='10' y='190' style = 'font-size: 80%;'>svg#radio2</text>
</svg>
</div>
```

<div style="margin-left: 30px">

```{js}
// movement function
var jump = function () {
      d3.select(this).transition().duration(500)
      .attr('y', '0')
      .transition().duration(500).ease(d3.easeBounce)
      .attr('y', '75');
};

// initial setup: add event listener to red square
d3.select("svg#radio2")
  .select("rect#red")
  .on("click", jump);
    
// switch event listeners if radio button is clicked
d3.select("div#rad2").selectAll("input")
  .on("click", function () {
  if (d3.select(this).node().value == "blue") {
    d3.select("svg#radio2").select("rect#blue").on("click", jump);
    d3.select("svg#radio2").select("rect#red").on("click", null);
    } else {
    d3.select("svg#radio2").select("rect#red").on("click", jump);
    d3.select("svg#radio2").select("rect#blue").on("click", null);
    }
});

```

</div>
