# Just Enough JS <i class="fab fa-js-square"></i>

<script src="https://d3js.org/d3.v7.js"></script>

Basics: *IDVW*, pp. 36-52

objects, arrays, arrays of objects, functions (and other things)

## Arrays of arrays

> <i class="fa fa-hand-o-right"></i> *Open the JavaScript Console* <i class="fa fa-terminal"></i>


``` js
// try me in the Console 
var array_dataset = [[100, 75, 30], [200, 125, 20]];

d3.select("svg#arrays")
  .selectAll("circle")
  .data(array_dataset)
  .enter()
  .append("circle")
    .attr("cx", d => d[0])
    .attr("cy", d => d[1])
    .attr("r", d => d[2])
    .attr("fill", "red");
```

```{r, echo=FALSE, results='asis'}
makesvg("arrays")
```


## Arrays of objects

``` js
// Try me in the Console
var object_dataset = [
  {cx: 100, cy: 150, fill: `red`},
  {cx: 200, cy: 100, fill: `blue`}
  ];

d3.select("svg#objects")
  .selectAll("circle")
  .data(object_dataset)
  .enter()
  .append("circle")
    .attr("cx", d => d.cx)
    .attr("cy", d => d.cy)
    .attr("r", "30")
    .attr("fill", d => d.fill);
```

```{r, results='asis', echo=FALSE}
makesvg("objects")
```

See also: [JavaScript Array of Objects Tutorial](https://www.freecodecamp.org/news/javascript-array-of-objects-tutorial-how-to-create-update-and-loop-through-objects-using-js-array-methods/)

## `.map()`

What's the issue?

In R many operations are [vectorized](https://bookdown.org/rdpeng/rprogdatascience/vectorized-operations.html){target="_blank"}:

```{r}
sqrt(3)
x <- c(3, 5, 7)
sqrt(x)
```

Not so in JavaScript:

``` js
Math.sqrt(3);     // Try me in the Console
```

``` js
var x = [3, 5, 7];     // Try me in the Console
Math.sqrt(x);          // Doesn't work...
```

### Simple arrays

Use `.map()` to operate on each array element separately. The concept is similar to `lapply()` or `purrr::map()`, but unlike in R, it's needed for simple arrays.

**R**

```{r}
x <- c(3, 5, 7)
sqrt(x)
```

**JavaScript**

Do something to every element of a simple array:

``` js
# take the square root of each element
var x = [3, 5, 7];     // try me
x.map(Math.sqrt);
```

``` js
# multiply each element by 3
[4, 10, 12].map(d => d*3);     // try me
```

``` js
# multiply each element by 3
[4, 10, 12].map(function(d) {return d*3;});     // try me
```

``` js
# multiply each element by its index
[10, 20, 30, 40].map((d, i) => d*i);     // try me
```

```{r}
# sum two arrays
x <- 1:3
y <- 4:6
x + y
```

``` js
# sum two arrays
var x = [1, 2, 3];
var y = [4, 5, 6];
x + y                         // try me... what went wrong?
```

``` js
# sum two arrays
var x = [1, 2, 3];
var y = [4, 5, 6];
x.map((d, i) => d + y[i]);     // try me
```


### Arrays of arrays

Do something to the first item of every element of a nested array:

``` js
[[1, 2], [3, 4]].map(d => Math.sqrt(d[0]))  // try me
```

Sum up all items in each element of the array:

``` js
[[1, 2, 3], [4, 5, 6]].map(d => d[0] + d[1] + d[2]); // try me
```

Created a nested array out of a simple array:

``` js
[10, 20, 30].map(d => [d, Math.pow(d, 2)]);
```

### Create arrays of objects

Create an array of objects out of a simple array (note the parentheses around the object):

``` js
[10, 20, 30].map(d => ({n: d, nsq: Math.pow(d, 2)}));  // try me
```

``` js
[10, 20, 30].map((d, i) => ({index: i, value: d}));  // try me
```

## D3 sorting

Use `d3.sort()` rather than plain JavaScript options.

``` js
var y = [3, 1, 5, 12, 7];     // try me
d3.sort(y);
```

## D3 statistics

[link to API](https://github.com/d3/d3-array/blob/v1.2.4/README.md#statistics){target="_blank"}

D3 brings us back to familiar ground with functions that take an *array* and return a single value. Here are D3 functions with the same names and behavior as their R equivalents:


|R|D3|
|:-|:-|
|`min(x)`|`d3.min(x)`|
|`max(x)`|`d3.max(x)`|
|`sum(x)`|`d3.sum(x)`|
|`mean(x)`|`d3.mean(x)`|
|`median(x)`|`d3.median(x)`|

<br>

A few with different names:

|R|D3|
|:-|:-|
|`range(x)`|`d3.extent(x)`|
|`var(x)`|`d3.variance(x)`|
|`sd(x)`|`d3.deviation(x)`|


<br>

`d3.quantile()` takes a single value for `p`, not an array as in R. (In earlier versions of D3 it was necessary to [sort the array](just-enough-js.html#sorting) before finding quantiles, but this is no longer the case.)


|R|D3|
|:-|:-|
|`quantile(x)`|`d3.quantile(x, p)`|

Thus for a single quantile we have:

``` js
var x = [12, 34, 1, 43, 90, 72];      // try me
d3.quantile(x, .25);
```

https://github.com/d3/d3/blob/master/API.md#statistics


## D3 + `.map()`

D3 statistics functions combined with `.map()` can be helpful in a variety of situations.

Vectorizing a parameter, for example to mimic `quantile(x)` in R:


**R**

```{r}
x <- c(1, 12, 34, 43, 72, 90);  
quantile(x)
```


**JavaScript**

``` js
var x = [1, 12, 34, 43, 72, 90];      // try me
[0, .25, .5, .75, 1].map(p => d3.quantile(x, p));
```


Sum up the first item of all elements in an array of arrays:

**R**

```{r}
l <- list(c(100, 200, 40), c(300, 150, 20))
sum(purrr::map_dbl(l, ~.x[1]))
```

**JavaScript**

``` js
var dataset = [[100, 200, 40], [300, 150, 20]];     // try me
d3.sum(dataset.map(d => d[0]));
```



Sum up all items in each array to create a simple array:

**R**

```{r}
l <- list(c(100, 200, 40), c(300, 150, 20))
purrr::map_dbl(l, ~sum(.x))
```

**JavaScript**

``` js
var dataset = [[100, 200, 40], [300, 150, 20]];     // try me
dataset.map(d => d3.sum(d));
```

